---
title: "Project  1"
author: "MMendoza"
format: 5
editor: visual
---

## Setup/Misc.

```{r}

# Uncomment Librarys Required
library(tidyverse)
library(readxl)
library(readr) 
library(DBI)
library(RSQLite)
library(tidycensus)
library(jsonlite)
library(httr)
library(lubridate)


# Set seed for random data creation for replication
set.seed(123)
```

## Part 1: Data Processing

### Parsing Variable Values

The function "variable_get" is a function designed to take the value of "YEAR" as a character input. It will return the variable list for the specified year ("2021", "2022", "2023", "2024") by creating the URL from the year associated with the .json file, parse the results into a datalist, and then will match and output the required variables for the specified year, and save the list as an .rds file in the "Data" subfolder.

```{r}
# Function to Parse Variable Values
variable_get <- function( YEAR) {
  
  # 1) Create jsonURL based on year selected
  jsonURL <- str_c("https://api.census.gov/data/", YEAR, 
                   "/acs/acs1/pums/variables.json")

  # 2) Use get() to call unparsed data
  unparsed_data <- httr::GET(jsonURL)
  str(unparsed_data, max.level = 1)
  
  # 3) parse returned information for the actual article information
  raw_json <- content(unparsed_data, as = "text")
  raw_data <- fromJSON(raw_json)
  parsed_data <- raw_data$variables
  
  # 4) Filter only necessary variables
    # 4.1) If  condition to alter needed variables based on year
    if (!(YEAR %in% c("2021", "2022", "2023", "2024"))) {
      stop("ERROR: input Variable YEAR is not in list of 2021, 2022, 2023, 2024")
    }
    else {
    if (YEAR %in% c("2021", "2022")) {
    varlist <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", 
                 "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", 
                 "REGION", "DIVISION", "ST")
    }
    else if (YEAR %in% c("2023", "2024")) {
    varlist <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                 "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", 
                 "REGION", "DIVISION", "STATE")
        
    }
    }
    # 4.2) create a list of variables that are both requested and present
    targets <- intersect(varlist, names(parsed_data))
    
    # 4.3) select matching vars from data. 
    output <- parsed_data[targets]
  
  
  # 5) save .RDS file to Datafolder
      # 5.1) Create unique file name per year to save to data folder
        outfile <- str_c("Data/", YEAR, ".rds")
        
      # 5.2) Save output as RDS
        saveRDS(output, file = outfile)
        
  #this is just for QC, print dataframe for review
  return(output)
}
```

The following code executes the "variable_get" function on the 4 requested years ("2021", "2022", "2023", "2024")

```{r}
#Call 2021 Data
variable_get("2021")

#Call 2022 Data
variable_get("2022")

#Call 2023 Data
variable_get("2023")

#Call 2024 Data
variable_get("2024")
```

### Parsing API Data

The code below is simply the set up requested in the outline. It specifies the URL for the year "2024" to show how to connect to the API database and return the unparsed results. this will be used to test helper function 1. the unparased data is stored as "unparsed_data2024".

```{r}
# Do the API get() once because the guidelines said to. 
# 1) assign URL to data
URL2024 <- "https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca"

# 2) Use get() to call unparsed data
unparsed_data2024 <- httr::GET(URL2024)
str(unparsed_data2024, max.level = 1)
```

The following is a helper function to parse the unparsed results from a GET() function. it uses the first row of the dataset to assign colnames, and all the rows after become the rows for the outputted tibble dataset.

```{r}
#HELPER FUNCTION 1: to parse and convert to tibble
parse_data <- function(unparsed_data){
   
#Content and prase the data
raw_json <- content(unparsed_data, as = "text")
raw_data <- fromJSON(raw_json)

# Store as tibble Handle data to clean and format columns
output <- as_tibble(raw_data[-1, ], .name_repair = "minimal") %>%
            set_names(toupper(raw_data[1, ])) 

#Return the results for outputing
return(output)
}
```

The following tests the helper function against the unparsed data from the first part of section "Parsing API Data". It stores the results as object "parsed_data_test" and prints the results

```{r}
parsed_data_test <- parse_data(unparsed_data2024)
print(parsed_data_test)
```

The following is a helper function. helped heavily by LLM to create, it takes a variable designated as "Time" variable and convert to a numeric datetime variable, adjusting for AM/PM as well as midnight transition. it then calculates the midpoint and returns the HMS time object. due to this being created mostly by a LLM, the testing will be done in the next function when it is used.

```{r}
#HELPER FUNCTION 2: DATE TIME (Helped from LLMs as per guidelines) time interval #strings and returns the midpoint time
parse_pums_time_midpoint <- function(time_code, var_name, all_levels) {
  # 1. Handle missing/invalid data inputs safely
  if (is.na(time_code) || time_code == "" || time_code == "N/A") return(hms(NA))
  
  # 2. Grab the actual descriptive string label mapped to the numeric code
  # Example: Code "020" maps to "06:00 a.m. to 06:29 a.m."
  time_str <- all_levels[[var_name]]$values$item[[as.character(time_code)]]
  
  if (is.null(time_str) || is.na(time_str)) return(hms(NA))
  
  # 3. Extract the two HH:MM timestamps with their AM/PM flags
  times <- str_extract_all(time_str, "\\d{2}:\\d{2}\\s*(a\\.m\\.|p\\.m\\.|am|pm)")[[1]]
  if (length(times) < 2) return(hms(NA)) 
  
  # 4. Extract hour/minute parts using standard lubridate structures
  t1 <- hm(str_extract(times[1], "\\d{2}:\\d{2}"))
  t2 <- hm(str_extract(times[2], "\\d{2}:\\d{2}"))
  
  # 5. Apply standardized 24-hour PM adjustments
  if (str_detect(tolower(times[1]), "p\\.m\\.|pm") && hour(t1) < 12) t1 <- t1 + hours(12)
  if (str_detect(tolower(times[2]), "p\\.m\\.|pm") && hour(t2) < 12) t2 <- t2 + hours(12)
  if (t2 < t1) t2 <- t2 + hours(24)
  
  # 6. Compute exact midpoint duration scale
  s1 <- as.numeric(t1)
  s2 <- as.numeric(t2)
  midpoint_seconds <- (s1 + s2) / 2
  
  return(seconds_to_period(midpoint_seconds %% 86400))
}

```

The following code creates a helper function that pulls the year .rds files to get the levels needed for specified categorical variables.

```{r}
getLevels <- function(YEAR){
  file_name <- paste0(YEAR, ".rds")
  file_path <- file.path("Data/", file_name)
  var_list <- readRDS(file_path)
  
  return(var_list)
}
```

This code test the helper function and verifies that the correct list can be pulled for a specified variable ("SEX").

```{r}
test <- getLevels(2021)
current_levels <- test[["ST"]]$values$item

current_levels

```

The following code creates the function "parse_pums". this function takes as input the specified year (2024 as default) in character format, the API key as a character, a vector of characters for the numeric variables ("AGEP", "PWGTP" are default), a vector of categorical variables ("SEX" as default), the geolevel required ("STATE" as default), and the specified geography subset specified (NULL by default to return the STATE info as needed).

```{r}
# Create Function to take year and key to parse data, convert to tibble, 
# and handle the various options requested in the project outline
parse_pums <- function(YEAR = "2024",
                       KEY = "2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca",
                       NUM_VARS = c("AGEP", "PWGTP"),
                       CAT_VARS = c("SEX"),
                       GEO_LEVEL = "STATE",
                       GEO_SUBSET = NULL){

# Verify Parameters are valid
# Numeric Variables
allowed_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
if (!all(NUM_VARS %in% allowed_numeric)) {
  stop("Error: Invalid numeric variable(s) specified.")
}
if (!"PWGTP" %in% NUM_VARS) {
  stop("Error: 'PWGTP' must always be included in NUM_VARS.")
}
if (length(setdiff(NUM_VARS, "PWGTP")) < 1) {
  stop("Error: You must specify at least one numeric variable other than 'PWGTP'.")
}

# Categorical Variables
allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
if (!all(CAT_VARS %in% allowed_categorical)) {
  stop("Error: Invalid categorical variable(s) specified.")
}
if (length(CAT_VARS) < 1) {
  stop("Error: You must specify at least one categorical variable.")
}

# Validate Geography Level
if (length(GEO_LEVEL) != 1 || GEO_LEVEL == "ALL" || !GEO_LEVEL %in% c("REGION", "DIVISION", "STATE")) {
  stop("Error: 'GEO_LEVEL' must be a single value of either 'REGION', 'DIVISION', or 'STATE'.")
}

# Save GEO_LEVEL before updating for default assignment
geo_trace <- GEO_LEVEL

# update GEO_LEVEL from STATE to ST based on year.
if (GEO_LEVEL == "STATE" && YEAR %in% c("2021", "2022")) {
  GEO_LEVEL <- "ST"
}

# Setup defaults if GEO_SUBSET is NULL
if (is.null(GEO_SUBSET) || !identical(GEO_SUBSET, "ALL")) {
  if (geo_trace == "STATE") {
    GEO_SUBSET <- "37" # Default to NC
  } else if (geo_trace == "DIVISION") {
    GEO_SUBSET <- "5"  # Default South Atlantic (South region)
  } else if (geo_trace == "REGION") {
    GEO_SUBSET <- "3"  # Default South Region
  }
}


# builds URL to connect to API and get unparsed data for specified Numeric
# and categorical variables.
base_url <- str_c("https://api.census.gov/data/", YEAR, "/acs/acs1/pums")
get_vars <- paste(unique(c(NUM_VARS, CAT_VARS)), collapse = ",")

# API Query to deal with ST/STATE discrepency
api_query <- if (geo_trace == "STATE") "state" 
             else tolower(GEO_LEVEL)

# Connect and pull data using a dynamic wildcard region request
res <- GET(base_url, query = list(
  get = get_vars, 
  `for` = str_c(api_query, ":*"), 
  key = KEY
))

# Use Helper 1 to parse data and create raw tibble
df <- parse_data(res)

# Subset tibble for Geographic parameters.
if (!identical(GEO_SUBSET, "ALL")) {
  df <- filter(df, .data[[geo_trace]] == GEO_SUBSET)
}

# use getLevels helper function to get list of variable levels. 
all_levels <- getLevels(YEAR)

# ---- 4. DATA PARSING & TYPE CONVERSIONS ----
# Implementation taken from LLM with minor adjustments.
# Parse Numeric and Time variables
for (var in NUM_VARS) {
  if (var %in% c("JWAP", "JWDP")) {
    # Dynamically map values by supplying the column code, column name, and levels      # metadata map
    df[[var]] <- map(df[[var]], ~parse_pums_time_midpoint(.x, var_name = var, all_levels = all_levels)) 
  } else {
    df[[var]] <- as.numeric(df[[var]])
  }
}

# Parse Categorical variables into factors
for (var in CAT_VARS) {
    # select current level from .rds List output from getLevels helper function
    current_levels <- all_levels[[var]]$values$item
    df[[var]] <- factor(df[[var]], levels = names(current_levels), 
                        labels = current_levels)
}

# parse GEO_LEVEL into factor variable
current_levels <- all_levels[[GEO_LEVEL]]$values$item
df[[geo_trace]] <- factor(df[[geo_trace]], levels = names(current_levels), 
                    labels = current_levels)

# do the last step to assign class to tibble. 
class(df) <- c("census", class(df))
return(df)
}
```

Test the function for various years and conditions

```{r}
#Test the function w/ Defaults only
data2024 <- parse_pums()

#Test the function w/ different YEAR
data2022 <- parse_pums(YEAR = "2022")

# Test the function for some everything changed (except APIKEY)
# This did work once but takes forever to run. so idk if it still does
# after some minor changes.
# data2022_sub <- parse_pums(YEAR = "2022",
#                        KEY = "2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca",
#                        NUM_VARS = c("AGEP", "PWGTP", "JWAP"),
#                        CAT_VARS = c("SEX", "FER"),
#                        GEO_LEVEL = "DIVISION",
#                        GEO_SUBSET = "ALL")

```

The following code creates a helper function to combine datasets and assigning the column of "YEAR" to the tibble.

```{r}
#Helper function to combine combine a list of single-year dataset
combine_pums_years <- function(data_list) {
  # Extract year tracking strings from the list names
  years <- names(data_list)
  
  # Inject the year column into each individual data frame
  for (i in seq_along(data_list)) {
    data_list[[i]] <- cbind(year = years[i], data_list[[i]])
  }
  
  # Row-bind all datasets cleanly using base R
  combined_df <- do.call(rbind, data_list)
  
  # Structure the final output as a census tibble
  final_tibble <- as_tibble(combined_df)
  class(final_tibble) <- c("census", class(final_tibble))
  
  return(final_tibble)
}

```

The following code wraps the list of variables and injects them into.

```{r}
#Function to excute all YEARS entered with specified values.
parse_pums_multi <- function(YEARS = c("2023", "2024"),
                             KEY = "2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca",
                             NUM_VARS = c("AGEP", "PWGTP"),
                             CAT_VARS = c("SEX"),
                             GEO_LEVEL = "STATE",
                             GEO_SUBSET = NULL) {
  
  # Pre-allocate our storage list
  data_list <- vector("list", length(YEARS))
  names(data_list) <- YEARS
  
  # Fetch data year-by-year
  for (i in seq_along(YEARS)) {
    data_list[[i]] <- parse_pums(
      YEAR = YEARS[i],
      KEY = KEY,
      NUM_VARS = NUM_VARS,
      CAT_VARS = CAT_VARS,
      GEO_LEVEL = GEO_LEVEL,
      GEO_SUBSET = GEO_SUBSET
    )
  }
  
  # Execute our separate combining function
  final_df <- combine_pums_years(data_list)
  return(final_df)
}

# test function
combined_data <- parse_pums_multi(CAT_VARS = c("SEX", "FER", "HHL"), NUM_VARS = c("AGEP", "PWGTP", "GASP"))
```

## Part 2: Function Writing

The following function creates a function to calculate Numeric Summaries for the census Summary method.

```{r}
#Summarize
summarize.census <- function(data, 
                             num_vars = NULL, 
                             cat_vars = NULL,
                             weight = "PWGTP") {
  
  # Default numeric variables: all numeric except PWGTP and time variables (if any)
  if (is.null(num_vars)) {
    num_vars <- data %>% 
      select(where(is.numeric)) %>% 
      select(-any_of(weight)) %>% 
      names()
  }
  
  # Default categorical variables: all non-numeric (or year treated as categorical)
  if (is.null(cat_vars)) {
    cat_vars <- data %>% 
      select(-any_of(c(num_vars, weight_col))) %>% 
      names()
  }
  
  results <- list()
  
  # Compute weighted means and standard deviations for numeric variables
  if (length(num_vars) > 0) {
    for (v in num_vars) {
      num_vec <- data[[v]]
      wt_vec  <- data[[weight]]
      
      # Sample mean formula provided
      s_mean <- sum(num_vec * wt_vec) / sum(wt_vec)
      # Sample standard deviation formula provided
      s_sd   <- sqrt(sum((num_vec^2) * wt_vec) / sum(wt_vec) - s_mean^2)
      
      results[[paste0(v, "_mean")]] <- s_mean
      results[[paste0(v, "_sd")]]   <- s_sd
    }
  }
  
  # Compute counts for categorical variables (unweighted or weighted depending on standard count, using n() or sum(PWGTP))
  if (length(cat_vars) > 0) {
    for (var in cat_vars) {
      cat_counts <- data %>%
        group_by(.data[[var]]) %>%
        summarize(count = n(), .groups = "drop")
      
      results[[paste0(var, "_counts")]] <- deframe(cat_counts)
    }
  }
  
  return(results)
}
```

The code below tests the function and prints the summary statistics from the combined year output from previous code.

```{r}
summary <- summary(combined_data)
summary
```

The following function creates a function that creates a generic plot function for a census class table.

```{r}
plot.census <- function(data,
                        cat_vars,
                        num_vars,
                        weight = "PWGTP") {
  # Validate Length of inputs
  if (!(length(cat_vars) = 1)) {
    stop("ERROR: cat_vars must be a length of 1")
  }
  if (!(length(cat_vars) = 1)) {
    stop("ERROR: num_vars must be a length of 1")
  }
  if (!(length(cat_vars) = 1)) {
    stop("ERROR: weight must be a length of 1")
  }
  
  plot_output <- ggplot(data, aes(x = get(cat_var), y = get(num_var), weight = get(weight)))
  
  #return plot object for output
  return(plot_output)
}

plot <- plot(data = combined_data, cat_vars = "FER", num_vars = "GASP")
plot
```
