---
title: "Project  1"
author: "MMendoza"
format: html
editor: visual
---

## Setup/Misc.

```{r}

# Uncomment Librarys Required
library(tidyverse)
library(readxl)
library(readr)
library(DBI)
library(RSQLite)
library(tidycensus)
library(jsonlite)
library(httr)
library(lubridate)


# Set seed for random data creation for replication
set.seed(123)
```

## Part 1: Data Processing

### Parsing Variable Values

The function "variable_get" is a function designed to take the value of "YEAR" as a character input. It will return the variable list for the specified year ("2021", "2022", "2023", "2024") by creating the URL from the year associated with the .json file, parse the results into a datalist, and then will match and output the required variables for the specified year, and save the list as an .rds file in the "Data" subfolder.

```{r}
# Function to Parse Variable Values
variable_get <- function( YEAR) {
  
  # 1) Create jsonURL based on year selected
  jsonURL <- str_c("https://api.census.gov/data/", YEAR, 
                   "/acs/acs1/pums/variables.json")

  # 2) Use get() to call unparsed data
  unparsed_data <- httr::GET(jsonURL)
  str(unparsed_data, max.level = 1)
  
  # 3) parse returned information for the actual article information
  raw_json <- content(unparsed_data, as = "text")
  raw_data <- fromJSON(raw_json)
  parsed_data <- raw_data$variables
  
  # 4) Filter only necessary variables
    # 4.1) If  condition to alter needed variables based on year
    if (!(YEAR %in% c("2021", "2022", "2023", "2024"))) {
      stop("ERROR: input Variable YEAR is not in list of 2021, 2022, 2023, 2024")
    }
    else {
    if (YEAR %in% c("2021", "2022")) {
    varlist <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", 
                 "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", 
                 "REGION", "DIVISION", "ST")
    }
    else if (YEAR %in% c("2023", "2024")) {
    varlist <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                 "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", 
                 "REGION", "DIVISION", "STATE")
        
    }
    }
    # 4.2) create a list of variables that are both requested and present
    targets <- intersect(varlist, names(parsed_data))
    
    # 4.3) select matching vars from data. 
    output <- parsed_data[targets]
  
  
  # 5) save .RDS file to Datafolder
      # 5.1) Create unique file name per year to save to data folder
        outfile <- str_c("Data/", YEAR, ".rds")
        
      # 5.2) Save output as RDS
        saveRDS(output, file = outfile)
        
  #this is just for QC, print dataframe for review
  return(output)
}
```

The following code executes the "variable_get" function on the 4 requested years ("2021", "2022", "2023", "2024")

```{r}
#Call 2021 Data
variable_get("2021")

#Call 2022 Data
variable_get("2022")

#Call 2023 Data
variable_get("2023")

#Call 2024 Data
variable_get("2024")
```

### Parsing API Data

The code below is simply the set up requested in the outline. It specifies the URL for the year "2024" to show how to connect to the API database and return the unparsed results. this will be used to test helper function 1. the unparased data is stored as "unparsed_data2024".

```{r}
# Do the API get() once because the guidelines said to. 
# 1) assign URL to data
URL2024 <- "https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca"

# 2) Use get() to call unparsed data
unparsed_data2024 <- httr::GET(URL2024)
str(unparsed_data2024, max.level = 1)
```

The following is a helper function to parse the unparsed results from a GET() function. it uses the first row of the dataset to assign colnames, and all the rows after become the rows for the outputted tibble dataset.

```{r}
#HELPER FUNCTION 1: to parse and convert to tibble
parse_data <- function(unparsed_data){
   
#Content and prase the data
raw_json <- content(unparsed_data, as = "text")
raw_data <- fromJSON(raw_json)

# Store as tibble Handle data to clean and format columns
output <- as_tibble(raw_data[-1, ], .name_repair = "minimal") %>%
            set_names(toupper(raw_data[1, ])) 

#Return the results for outputing
return(output)
}
```

The following tests the helper function against the unparsed data from the first part of section "Parsing API Data". It stores the results as object "parsed_data_test" and prints the results

```{r}
parsed_data_test <- parse_data(unparsed_data2024)
print(parsed_data_test)
```

The following is a helper function. helped heavily by LLM to create, it takes a variable designated as "Time" variable and convert to a numeric datetime variable, adjusting for AM/PM as well as midnight transition. it then calculates the midpoint and returns the HMS time object. due to this being created mostly by a LLM, the testing will be done in the next function when it is used.

```{r}
#HELPER FUNCTION 2: DATE TIME (Helped from LLMs as per guidelines) time interval #strings and returns the midpoint time
parse_pums_time_midpoint <- function(time_str) {
  if (is.na(time_str) || time_str == "" || time_str == "N/A") return(hms(NA))
  
  # Example pattern: "08:00 a.m. to 08:04 a.m."
  # Extract all HH:MM matches and AM/PM indicators
  times <- str_extract_all(time_str, "\\d{2}:\\d{2}\\s*(a\\.m\\.|p\\.m\\.|am|pm)")[[1]]
  
  if (length(times) < 2) return(hms(NA)) # Return NA if it doesn't match an interval
  
  # Convert extracted strings to full datetime objects to calculate midpoint cleanly
  t1 <- hm(str_extract(times[1], "\\d{2}:\\d{2}"))
  t2 <- hm(str_extract(times[2], "\\d{2}:\\d{2}"))
  
  # Adjust for PM
  if (str_detect(tolower(times[1]), "p\\.m\\.|pm") && hour(t1) < 12) t1 <- t1 + hours(12)
  if (str_detect(tolower(times[2]), "p\\.m\\.|pm") && hour(t2) < 12) t2 <- t2 + hours(12)
  # Handle midnight transitions if necessary (e.g., 11:50 pm to 12:05 am)
  if (t2 < t1) t2 <- t2 + hours(24)
  
  # Calculate midpoint seconds from midnight
  s1 <- as.numeric(t1)
  s2 <- as.numeric(t2)
  midpoint_seconds <- (s1 + s2) / 2
  
  # Wrap back into an hms time object (modulus handles rolling over 24 hours)
  return(seconds_to_period(midpoint_seconds %% 86400))
}

```

The following code creates a helper function that pulls the year .rds files to get the levels needed for specified categorical variables.

```{r}
getLevels <- function(YEAR){
  file_name <- paste0(YEAR, ".rds")
  file_path <- file.path("Data/", file_name)
  var_list <- readRDS(file_path)
  
  return(var_list)
}
```

The following code creates the function "parse_pums". this function takes as input the specified year (2024 as default) in character format, the API key as a character, a vector of characters for the numeric variables ("AGEP", "PWGTP" are default), a vector of categorical variables ("SEX" as default), the geolevel required ("STATE" as default), and the specified geography subset specified (NULL by default to return the STATE info as needed).

```{r}
# Create Function to take year and key to parse data, convert to tibble, 
# and handle the various options requested in the project outline
parse_pums <- function(YEAR = "2024",
                       KEY = "2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca",
                       NUM_VARS = c("AGEP", "PWGTP"),
                       CAT_VARS = c("SEX"),
                       GEO_LEVEL = "STATE",
                       GEO_SUBSET = NULL){

# Verify Parameters are valid
# Numeric Variables
allowed_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
if (!all(NUM_VARS %in% allowed_numeric)) {
  stop("Error: Invalid numeric variable(s) specified.")
}
if (!"PWGTP" %in% NUM_VARS) {
  stop("Error: 'PWGTP' must always be included in NUM_VARS.")
}
if (length(setdiff(NUM_VARS, "PWGTP")) < 1) {
  stop("Error: You must specify at least one numeric variable other than 'PWGTP'.")
}

# Categorical Variables
allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
if (!all(CAT_VARS %in% allowed_categorical)) {
  stop("Error: Invalid categorical variable(s) specified.")
}
if (length(CAT_VARS) < 1) {
  stop("Error: You must specify at least one categorical variable.")
}

# Validate Geography Level
if (length(GEO_LEVEL) != 1 || GEO_LEVEL == "ALL" || !GEO_LEVEL %in% c("REGION", "DIVISION", "STATE", "ST")) {
  stop("Error: 'GEO_LEVEL' must be a single value of either 'REGION', 'DIVISION', or 'STATE'.")
}

# ---- 2. HANDLE GEOGRAPHY SUBSETS & DEFAULTS ----
# Setup defaults if GEO_SUBSET is NULL
if (is.null(GEO_SUBSET)) {
  GEO_SUBSET <- switch(GEO_LEVEL,
                       "STATE" = "42",    # Default to Pennsylvania (42)
                       "DIVISION" = "2",  # Default Middle Atlantic Division
                       "REGION" = "1")    # Default Northeast Region
}

# Map user choice to the Census URL parameters
geo_param <- switch(GEO_LEVEL,
                    "STATE" = paste0("state:", 
                                     if(tolower(GEO_SUBSET) == "ALL") "*"
                                     else GEO_SUBSET),
                    "DIVISION" = paste0("division:", 
                                        if(tolower(GEO_SUBSET) == "ALL") "*" 
                                        else GEO_SUBSET),
                    "REGION" = paste0("region:", 
                                      if(tolower(GEO_SUBSET) == "ALL") "*" 
                                      else GEO_SUBSET)
                    )

# ---- 3. CONSTRUCT URL & GET DATA ----
base_url <- str_c("https://api.census.gov/data/", YEAR, "/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=")
get_vars <- paste(unique(c(NUM_VARS, CAT_VARS)), collapse = ",")

res <- GET(base_url, query = list(get = get_vars, `for` = geo_param, key = KEY))

# Use Helper 1 to create raw tibble
df <- parse_data(res)

# ---- 4. DATA PARSING & TYPE CONVERSIONS ----
# Parse Numeric and Time variables
for (var in NUM_VARS) {
  if (var %in% c("JWAP", "JWDP")) {
    # Apply Midpoint Time Helper (Helper 2)
    df[[var]] <- map(df[[var]], parse_pums_time_midpoint) 
  } else {
    df[[var]] <- as.numeric(df[[var]])
  }
}

#call the .rds list to parse categorical variables into factors. 
file_name <- paste0(YEAR, ".rds")
file_path <- file.path("Data/", file_name)
var_list <- readRDS(file_path)

# Parse Categorical variables into factors
# NOTE: Replace these mock lists with your readRDS("your_file.rds") level lookups!
levels <- test(YEAR)

for (var in CAT_VARS) {
    # select current level from .rds List output from getLevels helper function
    current_levels <- all_levels[[var]]
    df[[var]] <- factor(df[[var]], levels = names(current_levels), 
                        labels = current_levels)
}

# ---- 5. ASSIGN CLASS SUBCLASS & RETURN ----
class(df) <- c("census", class(df))
return(df)
}



#Test the function
data2023 <- parse_pums(2023)

```

```{r}


getLevels <- function(YEAR){
file_name <- paste0(YEAR, ".rds")
file_path <- file.path("Data/", file_name)
var_list <- readRDS(file_path)

return(var_list)
}

output_test <- getLevels("2024")






```

## Part 2: Function Writing
