---
title: "Project  1"
author: "MMendoza"
format: html
editor: visual
---

# Project 1

## Introduction

The following document is a program designed to call upon the census data available from "[api.census.gov](https://api.census.gov)". It will then create a function that can be used to pull in various types of variables from the API dataset. It will then handle factorize categorical variables, as well as create numeric time values of the mid point for specified time variables. It also creates a function that will combine a series of years if specified, and combine the data set into a single data table, adding the data source year as an additional column. Then a summary function is created that can be used to conduct the analysis at the end of the program. A generic plot function was also created to analyze whatever data set from the previous function we wanted to create.

The analysis that was conducted at the end of the this document, was an investigative analysis into the census data of

## Setup/Misc.

the following are the libraries that are or may be used in the program. as well as setting a randomization seed as part of good programming practices.

```{r}

# Uncomment Librarys Required
library(tidyverse)
library(readxl)
library(readr) 
library(DBI)
library(RSQLite)
library(tidycensus)
library(jsonlite)
library(httr)
library(lubridate)
library(ggplot2)
library(quantreg) #needed for ggplot2 to do weighted graphs


# Set seed for random data creation for replication
set.seed(123)
```

## Part 1: Data Processing

### Parsing Variable Values

The function "variable_get" is a function designed to take the value of "YEAR" as a character input. It will return the variable list for the specified year ("2021", "2022", "2023", "2024") by creating the URL from the year associated with the .json file, parse the results into a datalist, and then will match and output the required variables for the specified year, and save the list as an .rds file in the "Data" subfolder.

```{r}
# Function to Parse Variable Values
variable_get <- function( YEAR) {
  
  # 1) Create jsonURL based on year selected
  jsonURL <- str_c("https://api.census.gov/data/", YEAR, 
                   "/acs/acs1/pums/variables.json")

  # 2) Use get() to call unparsed data
  unparsed_data <- httr::GET(jsonURL)
  str(unparsed_data, max.level = 1)
  
  # 3) parse returned information for the actual article information
  raw_json <- content(unparsed_data, as = "text")
  raw_data <- fromJSON(raw_json)
  parsed_data <- raw_data$variables
  
  # 4) Filter only necessary variables
    # 4.1) If  condition to alter needed variables based on year
    if (!(YEAR %in% c("2021", "2022", "2023", "2024"))) {
      stop("ERROR: input Variable YEAR is not in list of 2021, 2022, 2023, 2024")
    }
    else {
    if (YEAR %in% c("2021", "2022")) {
    varlist <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", 
                 "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", 
                 "REGION", "DIVISION", "ST")
    }
    else if (YEAR %in% c("2023", "2024")) {
    varlist <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                 "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", 
                 "REGION", "DIVISION", "STATE")
        
    }
    }
    # 4.2) create a list of variables that are both requested and present
    targets <- intersect(varlist, names(parsed_data))
    
    # 4.3) select matching vars from data. 
    output <- parsed_data[targets]
  
  
  # 5) save .RDS file to Datafolder
      # 5.1) Create unique file name per year to save to data folder
        outfile <- str_c("Data/", YEAR, ".rds")
        
      # 5.2) Save output as RDS
        saveRDS(output, file = outfile)
        
  #this is just for QC, print dataframe for review
  return(output)
}
```

The following code executes the "variable_get" function on the 4 requested years ("2021", "2022", "2023", "2024") and saves the result to the Data subfolder.

```{r}
#Call 2021 Data
variable_get("2021")

#Call 2022 Data
variable_get("2022")

#Call 2023 Data
variable_get("2023")

#Call 2024 Data
variable_get("2024")
```

### Parsing API Data

The code below is simply the set up requested in the outline. It specifies the URL for the year "2024" to show how to connect to the API database and return the unparsed results. this will be used to test helper function 1. the unparased data is stored as "unparsed_data2024".

```{r}
# Do the API get() once because the guidelines said to. 
# 1) assign URL to data
URL2024 <- "https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca"

# 2) Use get() to call unparsed data
unparsed_data2024 <- httr::GET(URL2024)
str(unparsed_data2024, max.level = 1)
```

The following is a helper function to parse the unparsed results from a GET() function. it uses the first row of the data set to assign column names, and all the rows after become the rows for the outputted tibble data set. it also outputs the column names in all caps for formatting consistency among variables.

```{r}
#HELPER FUNCTION 1: to parse and convert to tibble
parse_data <- function(unparsed_data){
   
#Content and prase the data
raw_json <- content(unparsed_data, as = "text")
raw_data <- fromJSON(raw_json)

# Store as tibble Handle data to clean and format columns
output <- as_tibble(raw_data[-1, ], .name_repair = "minimal") %>%
            set_names(toupper(raw_data[1, ])) 

#Return the results for outputing
return(output)
}
```

The following tests the helper function against the unparsed data from the first part of section "Parsing API Data". It stores the results as object "parsed_data_test" and prints the results. As we can see, we get a tibble consisting of 47,476 rows and 5 variables included in the URL ("SEX", "PWGTP", "MAR", "STATE", and "SCHL"). All values are returned as characters.

```{r}
parsed_data_test <- parse_data(unparsed_data2024)
print(parsed_data_test)
```

The following is a helper function. helped heavily by LLM to create, it takes a variable designated as "Time" variable and convert to a numeric datetime variable, adjusting for AM/PM as well as midnight transition. it then calculates the midpoint and returns the HMS time object. due to this being created mostly by a LLM, the testing will be done in the next function when it is used. This helper function seems to have some difficulties with the data and only returns some times. not sure exactly why and I ran out of time trying to figure it out. but it does work sometimes :(.

```{r}
#HELPER FUNCTION 2: DATE TIME (Helped from LLMs as per guidelines) time interval #strings and returns the midpoint time
parse_pums_time_midpoint <- function(time_code, var_name, all_levels) {
  # 1. Handle missing/invalid data inputs safely
  if (is.na(time_code) || time_code == "" || time_code == "N/A") return(hms(NA))
  
  # 2. Grab the actual descriptive string label mapped to the numeric code
  # Example: Code "020" maps to "06:00 a.m. to 06:29 a.m."
  time_str <- all_levels[[var_name]]$values$item[[as.character(time_code)]]
  
  if (is.null(time_str) || is.na(time_str)) return(hms(NA))
  
  # 3. Extract the two HH:MM timestamps with their AM/PM flags
  times <- str_extract_all(time_str, "\\d{2}:\\d{2}\\s*(a\\.m\\.|p\\.m\\.|am|pm)")[[1]]
  if (length(times) < 2) return(hms(NA)) 
  
  # 4. Extract hour/minute parts using standard lubridate structures
  t1 <- hm(str_extract(times[1], "\\d{2}:\\d{2}"))
  t2 <- hm(str_extract(times[2], "\\d{2}:\\d{2}"))
  
  # 5. Apply standardized 24-hour PM adjustments
  if (str_detect(tolower(times[1]), "p\\.m\\.|pm") && hour(t1) < 12) t1 <- t1 + hours(12)
  if (str_detect(tolower(times[2]), "p\\.m\\.|pm") && hour(t2) < 12) t2 <- t2 + hours(12)
  if (t2 < t1) t2 <- t2 + hours(24)
  
  # 6. Compute exact midpoint duration scale
  s1 <- as.numeric(t1)
  s2 <- as.numeric(t2)
  midpoint_seconds <- (s1 + s2) / 2
  
  return(seconds_to_period(midpoint_seconds %% 86400))
}

```

The following code creates a helper function that pulls the year .rds files to get the levels needed for specified categorical variables. it calls upon the .rds list we made earlier, and returns a list of the levels needed for formatting categorical columns

```{r}
getLevels <- function(YEAR){
  file_name <- paste0(YEAR, ".rds")
  file_path <- file.path("Data/", file_name)
  var_list <- readRDS(file_path)
  
  return(var_list)
}
```

This code test the helper function and verifies that the correct list can be pulled for a specified variable ("ST" for example). returned is a list of all the variable values associated with the numeric code from the database.

```{r}
test <- getLevels(2021)
current_levels <- test[["ST"]]$values$item

current_levels

```

The following code creates the function "parse_pums". this function takes as input the specified year (2024 as default) in character format, the API key as a character, a vector of characters for the numeric variables ("AGEP", "PWGTP" are default), a vector of categorical variables ("SEX" as default), the geolevel required ("STATE" as default), and the specified geography subset specified (NULL by default to return the STATE info as needed). it includes various validation checks such as the following:

1.  Only permits the years for 2021, 2022, 2023, and 2024.
2.  Only allowing specified numeric variables.
3.  Allowed categorical variables.
4.  Verifies that required numeric variable "PWGTP" is included.
5.  Verifies at least "PWGTP" and 1 other numeric variable has been included.
6.  At least 1 categorical variable has been included.
7.  Verifies GEO_LEVEL is only 1 value of either REGION, DIVISION, and STATE
8.  Permits the use of GEO_SUBSET to include "ALL" so that all values of a GEO_LEVEL are included

The function also calls upon helper function 1 "parse_data" to parse the raw input data from the specified year of data. It calls upon helper function 2 "" to format "JWAP" and "JWDP" into numeric mid point values. it will finally assign factor levels to categorical variables and assign the class of census to the data frame.

```{r}
# Create Function to take year and key to parse data, convert to tibble, 
# and handle the various options requested in the project outline
parse_pums <- function(YEAR = "2024",
                       KEY = "2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca",
                       NUM_VARS = c("AGEP", "PWGTP"),
                       CAT_VARS = c("SEX"),
                       GEO_LEVEL = "STATE",
                       GEO_SUBSET = NULL){

# Verify Parameters are valid
# valid year
allowed_year <- c("2021", "2022", "2023", "2024")
if (!(YEAR %in% allowed_year)) {
  stop("Error: Invalid numeric year.")
}
  
# Numeric Variables
allowed_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
if (!all(NUM_VARS %in% allowed_numeric)) {
  stop("Error: Invalid numeric variable(s) specified.")
}
if (!"PWGTP" %in% NUM_VARS) {
  stop("Error: 'PWGTP' must always be included in NUM_VARS.")
}
if (length(setdiff(NUM_VARS, "PWGTP")) < 1) {
  stop("Error: You must specify at least one numeric variable other than 'PWGTP'.")
}

# Categorical Variables
allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")
if (!all(CAT_VARS %in% allowed_categorical)) {
  stop("Error: Invalid categorical variable(s) specified.")
}
if (length(CAT_VARS) < 1) {
  stop("Error: You must specify at least one categorical variable.")
}

# Validate Geography Level
if (length(GEO_LEVEL) != 1 || GEO_LEVEL == "ALL" || !GEO_LEVEL %in% c("REGION", "DIVISION", "STATE")) {
  stop("Error: 'GEO_LEVEL' must be a single value of either 'REGION', 'DIVISION', or 'STATE'.")
}

# Save GEO_LEVEL before updating for default assignment
geo_trace <- GEO_LEVEL

# update GEO_LEVEL from STATE to ST based on year.
if (GEO_LEVEL == "STATE" && YEAR %in% c("2021", "2022")) {
  GEO_LEVEL <- "ST"
}

# Setup defaults if GEO_SUBSET is NULL
if (is.null(GEO_SUBSET) || !identical(GEO_SUBSET, "ALL")) {
  if (geo_trace == "STATE") {
    GEO_SUBSET <- "37" # Default to NC
  } else if (geo_trace == "DIVISION") {
    GEO_SUBSET <- "5"  # Default South Atlantic (South region)
  } else if (geo_trace == "REGION") {
    GEO_SUBSET <- "3"  # Default South Region
  }
}


# builds URL to connect to API and get unparsed data for specified Numeric
# and categorical variables.
base_url <- str_c("https://api.census.gov/data/", YEAR, "/acs/acs1/pums")
get_vars <- paste(unique(c(NUM_VARS, CAT_VARS)), collapse = ",")

# API Query to deal with ST/STATE discrepency
api_query <- if (geo_trace == "STATE") "state" 
             else tolower(GEO_LEVEL)

# Connect and pull data using a dynamic wildcard region request
res <- GET(base_url, query = list(
  get = get_vars, 
  `for` = str_c(api_query, ":*"), 
  key = KEY
))

# Use Helper 1 to parse data and create raw tibble
df <- parse_data(res)

# Subset tibble for Geographic parameters.
if (!identical(GEO_SUBSET, "ALL")) {
  df <- filter(df, .data[[geo_trace]] == GEO_SUBSET)
}

# use getLevels helper function to get list of variable levels. 
all_levels <- getLevels(YEAR)

# ---- 4. DATA PARSING & TYPE CONVERSIONS ----
# Implementation taken from LLM with minor adjustments.
# Parse Numeric and Time variables
for (var in NUM_VARS) {
  if (var %in% c("JWAP", "JWDP")) {
    # Dynamically map values by supplying the column code, column name, and levels      # metadata map
    df[[var]] <- map(df[[var]], ~parse_pums_time_midpoint(.x, var_name = var, all_levels = all_levels)) 
  } else {
    df[[var]] <- as.numeric(df[[var]])
  }
}

# Parse Categorical variables into factors
for (var in CAT_VARS) {
    # select current level from .rds List output from getLevels helper function
    current_levels <- all_levels[[var]]$values$item
    df[[var]] <- factor(df[[var]], levels = names(current_levels), 
                        labels = current_levels)
}

# parse GEO_LEVEL into factor variable
current_levels <- all_levels[[GEO_LEVEL]]$values$item
df[[geo_trace]] <- factor(df[[geo_trace]], levels = names(current_levels), 
                    labels = current_levels)

# do the last step to assign class to tibble. 
class(df) <- c("census", class(df))
return(df)
}
```

the following test the function for various years and conditions that SHOULD work. The final line of code has been commented out due to long processing times or possible error causing an infinite feed back loop.

```{r}
#Test the function w/ Defaults only
data2024 <- parse_pums()
# output looks good! we end up with a dataset of 4 variables, AGEP as a numeric 
# variabled, PWGTP as a numeric variable, SEX as a factor variable with 2 levels 
# of "Male" and "Female", and STATE as a factor variable with 52 levels.

# Test the function for some everything changed (except APIKEY)
 data2022_sub <- parse_pums(YEAR = "2022",
                        KEY = "2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca",
                        NUM_VARS = c("AGEP", "PWGTP", "JWAP"),
                        CAT_VARS = c("SEX", "FER"),
                        GEO_LEVEL = "DIVISION",
                        GEO_SUBSET = "1")
 # output mostly looks good, the added variables of JWAP and FER are included in
 # addion to the other variables we specified (same as the defaults). One
 # issue we can immidately notice is that PWGTP is sometimes null. I'm not exactly
 # sure why this is, other than one of the conditions in the helper function
 # doesn't account for some situation in the source data. This is definitely 
 # fixable but I'd need more time to change the function so I can see the raw 
 # output from the dataset, as well as updating the helper function to output
 # all objects to R so I can review what part of the code is malfunctioning.or
 # maybe there's just that many NAs. It just seems odd.

 # Test the function for GEO_SUBSET = "ALL"
 # I've excluded this code because I don't fully know if it works or not. 
 # because I could not get the all function working appriopately.
 # my idea was that ALL would include the 3 other variables
 # when calling the API and include them as all :* in the url. 
 # however I could not figure out exactly how to format this without my code 
 # breaking. there should be some way to accomplish it by this method. 
 # data2022_sub <- parse_pums(GEO_SUBSET = "ALL")
```

The following code intentionally tries to fail the code to verify the validation checks are functioning.

```{r}
#| eval: false
# year check failure
failed <- parse_pums(YEAR = "2000")
# validation success!
```

```{r}
#| eval: false
# Numeric check failures
failed <- parse_pums(NUM_VARS = c("PWGTP", "FAKE"))
# Validition success!
```

```{r}
#| eval: false
# Categorical check failures
failed <- parse_pums(CAT_VARS = c("FAKE"))
# Validition success!
```

```{r}
#| eval: false
# PWGTP not included check failures
failed <- parse_pums(NUM_VARS = c("AGEP"))
# Validition success!
```

```{r}
#| eval: false
# PWGTP and 1 other Numeric variable check failures
failed <- parse_pums(NUM_VARS = c("PWGTP"))
# Validition success!
```

```{r}
#| eval: false
# at least 1 categorical varaible check failures
failed <- parse_pums(CAT_VARS = NULL)
# Validition success!
```

```{r}
#| eval: false
# GEO_LEVEL check failures: 1 value not in Region, State, or Division. 
# ALso that "ALL" doesn't work is sort of implicit by the first condition of the
# 3 permitted values. however it was included as requested.
failed <- parse_pums(GEO_LEVEL = c("PWGTP"))
# Validition success!
```

```{r}
#| eval: false
# GEO_LEVEL check failures: Multiple inputs
failed <- parse_pums(GEO_LEVEL = c("STATE", "REGION"))
# Validition success!
```

The following code creates a helper function to combine datasets and assigning the column of "YEAR" to the tibble. it accomplishes this by assigning the year to each variable year within a datalist. it then combines the datalist together to output a single table. this datalist wil be created in the next function called "parse_pums_multi".

```{r}
# Helper function to combine combine a list of single-year dataset.
combine_pums_years <- function(data_list) {
  # Extract year tracking strings from the list names
  years <- names(data_list)
  
  # Inject the year column into each individual data frame
  for (i in seq_along(data_list)) {
    data_list[[i]] <- cbind(year = years[i], data_list[[i]])
  }
  
  # Row-bind all datasets
  combined_df <- do.call(rbind, data_list)
  
  # Structure the final output as a census tibble
  final_tibble <- as_tibble(combined_df)
  class(final_tibble) <- c("census", class(final_tibble))
  
  return(final_tibble)
}

```

The following code executes the parse_pums function for each year, where it will store each year separate inside of a data list, and then will use the "combine_pums_years" function to create and output a single tibble, with an additional column added for each year the data set originates from.

```{r}
# Function to excute all YEARS entered with specified values.
parse_pums_multi <- function(YEARS = c("2023", "2024"),
                             KEY = "2aa23f6929b5b51bb7e24e90cc25a4c02d9290ca",
                             NUM_VARS = c("AGEP", "PWGTP"),
                             CAT_VARS = c("SEX"),
                             GEO_LEVEL = "STATE",
                             GEO_SUBSET = NULL) {
  
  # Pre-allocate our storage list
  data_list <- vector("list", length(YEARS))
  names(data_list) <- YEARS
  
  # Fetch data year-by-year
  for (i in seq_along(YEARS)) {
    data_list[[i]] <- parse_pums(
      YEAR = YEARS[i],
      KEY = KEY,
      NUM_VARS = NUM_VARS,
      CAT_VARS = CAT_VARS,
      GEO_LEVEL = GEO_LEVEL,
      GEO_SUBSET = GEO_SUBSET
    )
  }
  
  # Execute our separate combining function
  final_df <- combine_pums_years(data_list)
  return(final_df)
}
```

The following code is an example of the parse_pums_multi function in use. for this, I simply used the default years of 2023 and 2024, with additional CAT_VARS variables that will be investigated during the next part of the project.

```{r}
# test function
combined_data <- parse_pums_multi(NUM_VARS = c("AGEP", "PWGTP", "GASP"),
                                  CAT_VARS = c("SEX", "FER", "HHL"))



```

## Part 2: Function Writing

The following function creates a function to calculate Numeric Summaries for the census Summary method. it first auto populates all numeric and categorical variables permited into num_vars and cat_vars when either is null. then it computes the weighted means and standard deviations based on the formulas provided for all numeric variables, Lastly, it computes the counts for each categorical variables and returns the results for review.

```{r}
# Summarize Function
summarize.census <- function(data, 
                             num_vars = NULL, 
                             cat_vars = NULL,
                             weight = "PWGTP") {
  
  # Selects all numeric variables except specified ones if num_vars is NULL
  if (is.null(num_vars)) {
    num_vars <- data %>% 
      select(where(is.numeric)) %>% 
      select(-any_of(weight)) %>% 
      names()
  }
  
  # Selects all categorical variables except specified ones if cat_vars is NULL
  if (is.null(cat_vars)) {
    cat_vars <- data %>% 
      select(-any_of(c(num_vars, weight_col))) %>% 
      names()
  }
  
  results <- list()
  
  # Compute weighted means and standard deviations for numeric variables
  if (length(num_vars) > 0) {
    for (var in num_vars) {
      num_vec <- data[[var]]
      wt_vec  <- data[[weight]]
      
      # Sample mean formula provided
      s_mean <- sum(num_vec * wt_vec) / sum(wt_vec)
      # Sample standard deviation formula provided
      s_sd   <- sqrt(sum((num_vec^2) * wt_vec) / sum(wt_vec) - s_mean^2)
      
      results[[paste0(v, "_mean")]] <- s_mean
      results[[paste0(v, "_sd")]]   <- s_sd
    }
  }
  
  # Compute counts for categorical variables
  if (length(cat_vars) > 0) {
    for (var in cat_vars) {
      cat_counts <- data %>%
        group_by(.data[[var]]) %>%
        summarize(count = n(), .groups = "drop")
      
      results[[paste0(var, "_counts")]] <- deframe(cat_counts)
    }
  }
  
  return(results)
}
```

The code below tests the function and prints the summary statistics from the combined year output from previous code. the output is a bit hard to read due to the long factor level names. otherwise results seem good.

```{r}
summary <- summary(combined_data)
summary
```

The following function creates a function that creates a generic plot function for a census class table. It first includes 3 checks to verify that cat_vars, num_vars, and weight are all exactly 1 variable. then, it uses ggplot2 and optional ggplot2 package "quantreg" to create a box plot of the weighted numeric variable by the categorical variable. it then returns the plot for it to be called as needed. colors are assigned by the category variable, weighted by specified weighted variable (PWGTP by default).

```{r}
# Create the plot.census function
plot.census <- function(data,
                        cat_vars,
                        num_vars,
                        weight = "PWGTP") {
  # Validate Length of inputs
  if (!(length(cat_vars) = 1)) {
    stop("ERROR: cat_vars must be a length of 1")
  }
  if (!(length(cat_vars) = 1)) {
    stop("ERROR: num_vars must be a length of 1")
  }
  if (!(length(cat_vars) = 1)) {
    stop("ERROR: weight must be a length of 1")
  }
  
  plot_output <- ggplot(data, aes(x = get(cat_vars), y = get(num_vars), 
                                  weight = get(weight), fill = get(cat_vars))) +
                    geom_boxplot() + 
                    labs(
                      x = cat_vars,
                      y = num_vars,
                      title = paste("Weighted Boxplot of", num_vars, 
                                    "by", cat_vars),
                      fill = cat_vars
                    )
  
  #return plot object for output
  return(plot_output)
}
```

The following code is an example use of the plot function, using the earlier combined data set made from the parse_pums_multi example test. The example shows the boxplot of the monthly gas cost compared relative to the groups of people who have had a child with the past 12 Months of the census ("N/A", "No", "Yes"). we can see the plot works, though this isn't very interesting as there's not much logic to how having a child and the cost of gas relate to one another.

```{r}
# test the plot.census function
plot <- plot(data = combined_data, cat_vars = "FER", num_vars = "GASP") 
plot
```

## Investigation

In order to demonstrate using these functions to investigate the database, a comparison of the different age among the different sexes within a the North Carolina state during the 2024. to do this, we first must use the parse_pums function. Thankfully, we can use the default settings for most of the function, but we will need to adjust the year to 2023.

```{r}
# call 2023 data for NC state as a tibble including only default variables
data2023 <- parse_pums(YEAR = "2023")
```

with our dataset now created, we can use summarize.census general function in order to get a brief overview of the dataset. This we can review and see if there are any abnormalities that indicate something wrong with out dataset.

```{r}
# execute summary function and display results
summary <- summary(data2023)
summary
```

With our summary results, we see that there are there are 54412 Males and 57508 Females. Among both of these sexes, we see that the oldest a person recorded is 93, and the youngest is 0, with 25% of all people being below the age of 22, 50% below the age of 44, 75% below the age of 63, and the average age being 43.43 years old.

Now use the plot.census function to see the boxplot for males and female ages, and see if we notice anything of note.

```{r}
# create boxplot of age vs sex using plot.census function
plot <- plot(data = data2023, cat_vars = "SEX", num_vars = "AGEP") 
plot
```

from this, we can see the medians, IQR, and Minimum/Maximum for both Male and Female. we can noticed that the IQR for females, seems to be slightly higher than that for Males, as well as the median also being slightly higher. that said, it seems that men and women live to be roughly the same age based on the visual of the graph.
